\subsection{Apache Calcite vision}
\label{subsec:vision}

In this paper we describe Apache Calcite, a data management framework that aims at meeting the aforementioned goals. The dynamic nature of Calcite comes from its ability to adapt to the requirements of the underlying processing platforms depending on their specific needs. Calcite has enjoyed wide adoption since its inception, and more than a dozen systems widely used in industry already use it to power their query optimization logic~\footnote{\url{http://calcite.apache.org/docs/powered_by}}.

In the following, we enumerate some of the major features of Calcite that have contributed to the wide adoption of the framework.

\begin{itemize}
	\item\textbf{Open source friendly.} Calcite is an open-source framework, backed by the Apache Software Foundation (ASF)~\cite{asf:website}, which provides the means to collaboratively develop the project. Furthermore, the software is written in Java, making it easy to interoperate with many of the latest developed data processing systems~\cite{website:Drill,website:Flink,website:Hive,website:Kylin,website:Samza}, especially those in the Hadoop ecosystem.
	\item\textbf{Multiple data models.} Support for optimization of streaming and non-streaming data processing paradigms integrated within the same abstraction layer.
	\item\textbf{Flexible query optimizer.} Each component of the optimizer is pluggable and extensible, ranging from rules to cost models. In addition, Calcite includes support for multiple planning engines, making it easy to construct a multi-phase optimization process.
	\item\textbf{Cross-platform support.} The framework can optimize queries whose execution span across multiple processing systems. Even better, the decision is seamlessly integrated within the optimization logic by representing the system for each operator as a simple physical property.
	\item\textbf{Reliable and efficient implementation.} Calcite is reliable, as its wide adoption has led to exhaustive testing of the platform. Furthermore, its implementation includes techniques to improve the efficiency of the query optimization process, \eg\ smart caching of metadata information.
	\item\textbf{Extensions to work out-of-the-box.} Many systems do not provide their own query language, but rather prefer to rely on existing ones. For those, Calcite provides support for standard SQL, as well as various SQL dialects and extensions, \eg\ for expressing queries on streaming or nested data. In addition, Calcite includes a JDBC driver that can be used to easily integrate with the optimization framework.
\end{itemize}
